---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Load the data.

```{r}
# Unzip and read the data file.
activity <- read.csv(unzip("activity.zip"))

```

Preproces the data.

- Create a new factor variable 'is weekend' in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
- For checking purposes, add a new variable 'day of week' with seven levels from Sunday to Saturday.

```{r}
# We want to see the day of the week in English: locale "C" reflects North-American usage.
Sys.setlocale(category = "LC_ALL", locale = "C")

# Create a factor variable 'is weekend' in the dataset with two levels - "weekday" and "weekend".
#   In English the day of the week starting with character S are Saturday and
#   Sunday - they mark the weekend (and the rest are weekdays).
#   The 'starting with' is notated with character circumflex accent (^).
activity$is_weekend <- as.factor(as.numeric(as.factor(grepl('^S', as.factor(weekdays(as.Date(activity$date)))))))
levels(activity$is_weekend) <- c("weekday", "weekend")

# For checking purposes only, add a new variable 'day of week' with seven levels from Sunday to Saturday.
activity$weekday <- as.factor(weekdays(as.Date(activity$date)))
# .. and order them starting from Sunday (the US way).
activity$weekday <- factor(activity$weekday, 
  levels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

###Result dataset
Let's have a look at our dataset 'activity'.

```{r}
str(activity)
```

A sample of the data .
```{r}
activity[9502:9508,]
```


## What is mean total number of steps taken per day?

**1. Calculate the total number of steps taken per day.**

```{r}
# With function aggregate, the default is to ignore missing values in the given variables.
steps_per_day <- aggregate(steps ~ date, data=activity, sum)
str(steps_per_day)

step_total <- sum(steps_per_day$steps)
step_total

```

The total number of steps taken per day is `r step_total`.



**2. Make a histogram of the total number of steps taken each day.**
```{r}
hist(steps_per_day$steps, breaks=12, xlab="Steps", main="Steps per day")
```

**3. Calculate and report the mean and median of the total number of steps taken per day.**

```{r}
step_mean <- mean(steps_per_day$steps)
step_mean

step_median <- median(steps_per_day$steps)
step_median
```

The measures of central tendency for steps taken are

 - the mean is `r format(step_mean,sigits=2)` and 
 - the median is `r step_median`.



## What is the average daily activity pattern?

**1. Make a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days.**

```{r}
# With function aggregate, the default is to ignore missing values in the given variables.
avg_steps_per_interval <- aggregate(steps ~ interval, data=activity, mean)
str(avg_steps_per_interval)
plot(avg_steps_per_interval, type="l", xlab="5-minute interval", main="Average daily activity")
interval_containing_max_steps <- avg_steps_per_interval[which.max(avg_steps_per_interval$steps), ]
max_text <- paste("The maximun steps is", as.character(sprintf("%.4f", interval_containing_max_steps$steps)), " at interval", as.character(interval_containing_max_steps$interval))
text(c(1600),c(206),labels=max_text)
```

**2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?**

```{r}
avg_steps_per_interval[which.max(avg_steps_per_interval$steps), ]
```
`r max_text`.

## Imputing missing values

**1. Calculate and report the total number of missing values in the dataset**
```{r}
NA_count <- sum(is.na(activity))
NA_count
```

There are `r NA_count` missing values.


**2. Devise a strategy for filling in all of the missing values in the dataset. **

Replace the NAs with '5-minute interval averages'
```{r}
activity_imp <- merge(activity, avg_steps_per_interval, by = "interval", suffixes = c("", ".na"))
# Create an index list of NAs
na_ind <- is.na(activity_imp$steps)
# Replace NAs
activity_imp$steps[na_ind] <- activity_imp$steps.na[na_ind]
```


**3. Create a new dataset that is equal to the original dataset but with the missing data filled in.**
```{r}
# Remove column 'steps.na' (the sixth colomn)
activity_imp <- activity_imp[, c(1:5)]
str(activity_imp)
```

**4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.**

```{r}
steps_per_day_imp <- aggregate(steps ~ date, data=activity_imp, sum)
hist(steps_per_day$steps, breaks=12, xlab="Steps", main="Steps per day (Imputing NAs")

step_mean_imp <- mean(steps_per_day_imp$steps)
step_mean_imp

step_median_imp <- median(steps_per_day_imp$steps)
step_median_imp

# steps before (with NAs)
step_total
# steps after (NAs replaces wit averages)
step_total_imp <- sum(steps_per_day_imp$steps)
step_total_imp

# difference is steps 
more_steps = (step_total_imp - step_total)
more_steps
str_more_steps = sprintf("%.0f", more_steps)

```


After imputing missing values, the measures of central tendency for steps taken are

 - the mean is `r format(step_mean_imp,sigits=2)` (*before `r format(step_mean,sigits=2)`*) and
 - the median is `r format(step_median_imp,sigits=2)` (*before `r format(step_median,sigits=2)`*).


The impact of imputing missing data **did not change the mean** and affected the **median only slightly**.

However, replacing the NAs added the total number of steps taken with `r str_more_steps` steps.


## Are there differences in activity patterns between weekdays and weekends?


**1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend"**

This was already done at the beginnig in the preprocessing phase; see variable 'is_weekend'. 

**2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days.**



```{r}
par(mfrow = c(2, 1))
for (is_we in c("weekend", "weekday")) {
    steps.is_we <- aggregate(steps ~ interval, data = activity_imp, subset = activity_imp$is_weekend == 
        is_we, FUN = mean)
    plot(steps.is_we, type = "l", main = is_we)
}

```

It seems that there is more activity on weekend than on weekdays. 
Even though the subject get's up later he/she is more active during the day (and goes to bed a maybe a bit later than on weekdays).
